Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/biometric_vectorization_example.tex
Minor tweaks to Extension with Array Writes. (Made more precise.)
Changed language=cppp to language=C++ in macros.

We start from Benjamin's code with linear loops (MPC Source):

{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
for i in range(0, N!0):
   min_sum!2 = PHI(min_sum!1, min_sum!4)
   min_index!2 = PHI(min_index!1, min_index!4)
   sum!2 = 0
   for j in range(0, D!0):
     sum!3 = PHI(sum!2, sum!4)
     d!3 = (S!0[((i * D!0) + j)] - C!0[j])
     p!3 = (d!3 * d!3)
     sum!4 = (sum!3 + p!3)
   !1!2 = (sum!3 < min_sum!2)
   min_sum!3 = sum!3
   min_index!3 = i
   min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2)
   min_index!4 = MUX(!1!2, min_index!3, min_index!2)
!2!1 = (min_sum!2, min_index!2)   
\end{verbatim}
}

@@

In the code below, all initializations (e.g., \texttt{min\_sum!3 = i}), operations, and PHI nodes are \emph{implicitly vectorized}. 
$\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ statements have slightly different interpretation. %\ana{This is vague, may need elaboration.}

The example illustrates the two different versions of $\mathit{raise\_dim}$. \texttt{C!0' = raise\_dim(C!0, j, (i:N!0,j:D!0))} reshapes
the read-only input array, while \texttt{sum!3' = drop\_dim(sum!3)} removes the $j$ dimension of \texttt{sum!3}.
%\ana{More here! One just adds a least-significant dimension. The other one may reshape an input array.}

{\small
\begin{verbatim}
min_sum!1 = 10000
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))
for i in range(0, N!0):
   min_sum!2 = PHI(min_sum!1^, min_sum!4)
   min_index!2 = PHI(min_index!1^, min_index!4) 
   sum!2 = 0 // Will lift, when hoisted
   sum!2^ = raise_dim(sum!2, (j:D!0)) // Special form?
   for j in range(0, D!0):
      sum!3 = PHI(sum!2^, sum!4)
      d!3 = S!0^ - C!0^
      p!3 = (d!3 * d!3) 
      sum!4 = (sum!3 + p!3)
   sum!3^ = drop_dim(sum!3)     
   !1!2 = (sum!3^ < min_sum!2)
   min_sum!3 = sum!3^
   min_index!3 = i  // Same-level, will lift when hoisted
   min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2)
   min_index!4 = MUX(!1!2, min_index!3, min_index!2)
min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)     
\end{verbatim}
}

@@

Starting at the inner phi-node \texttt{sum!3 = PHI(sum!2\^, sum!4)}, the algorithm first computes its closure. The closure amounts to the phi-node itself and the addition node \texttt{sum!4 = (sum!3 + p!3)}, accounting for the loop-carried dependency of the computation of \texttt{sum}. The algorithm replaces this closure with a FOR loop on $j$ removing vectoriaztion on $j$. Note that the SUB and MUL computations remain outside of the loop as they do not depend on phi-nodes that are part of cycles. The dependences are from \texttt{p!3[I,J] = (d!3[I,J] * d!3[I,J])} to the monolithic FOR loop and from the FOR loop to \texttt{sum!3\^ = drop\_dim(sum!3)}. (Lower case index, e.g., \texttt{i}, indicates non-vectorized dimension, while uppercase index, e.g., \texttt{I} indicates vectorized dimension.)

@@

{\small
\begin{verbatim}
min_sum!1 = 10000
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))
for i in range(0, N!0):
  min_sum!2[I] = PHI(min_sum!1^[I], min_sum!4[I]) 
  min_index!2[I] = PHI(min_index!1^[I], min_index!4[I])  
  sum!2 = [0,..,0] 
  sum!2^ = raise_dim(sum!2, (j:D!0))
  d!3[I,J] = S!0^[I,J] - C!0^[I,J]
  p!3[I,J] = (d!3[I,J] * d!3[I,J])
  for j in range(0, D!0):
    sum!3[I,j] = PHI(sum!2^[I,j], sum!4[I,j-1])       
    sum!4[I,j] = (sum!3[I,j] + p!3[I,j])
  sum!3^ = drop_dim(sum!3)     
  !1!2[I] = (sum!3^[I] < min_sum!2[I])
  min_sum!3 = sum!3^
  min_index!3 = i
  min_sum!4[I] = MUX(!1!2[I], min_sum!3[I], min_sum!2[I])
   min_index!4[I] = MUX(!1!2[I], min_index!3[I], min_index!2[I])
min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)
\end{verbatim}
}

When processing the outer loop two closures arise, one for \texttt{min\_sum!2[I] = PHI(...)} and one 
for \texttt{min\_index!2[I] = PHI(...)}. Since the two closures \emph{do not} intersect, we have two distinct FOR-loops on $i$:

{\small
\begin{verbatim}
min_sum!1 = 10000
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)
\end{verbatim}
}

@@

{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

// drop_dim becomes redundant
!2!1 = (min_sum!2, min_index!2)
\end{verbatim}
}

@@

{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

sum!2 = [0,..,0]
d!3[I,J] = SUB_SIMD(S!0^[I,J],C!0^[I,J])
p!3[I,J] = MUL_SIMD(d!3[I,J] * d!3[I,J])

@@

!2!1 = (min_sum!2, min_index!2)   
\end{verbatim}
}

>>>>

We start from MPC Source:

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_index!1 = 0
for i in range(0, N!0):
   min_sum!2 = PHI(min_sum!1, min_sum!4)
   min_index!2 = PHI(min_index!1, min_index!4)
   sum!2 = 0
   for j in range(0, D!0):
     sum!3 = PHI(sum!2, sum!4)
     d!3 = (S!0[((i * D!0) + j)] - C!0[j])
     p!3 = (d!3 * d!3)
     sum!4 = (sum!3 + p!3)
   !1!2 = (sum!3 < min_sum!2)
   min_sum!3 = sum!3
   min_index!3 = i
   min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2)
   min_index!4 = MUX(!1!2, min_index!3, min_index!2)
!2!1 = (min_sum!2, min_index!2)   
\end{pythonn}
}

@@

In the code below statements (e.g., {\sf min\_sum!3 = i}) are implicitly vectorized. 
%$\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ statements have slightly different interpretation. %\ana{This is vague, may need elaboration.}
The example illustrates the two different versions of $\mathit{raise\_dim}$. {\sf raise\_dim(C!0, j, (i:N!0,j:D!0))} reshapes
the read-only input array, while {\sf drop\_dim(sum!3)} removes the $j$ dimension of {\sf sum!3}.
%\ana{More here! One just adds a least-significant dimension. The other one may reshape an input array.}

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))
for i in range(0, N!0):
   min_sum!2 = PHI(min_sum!1^, min_sum!4)
   min_index!2 = PHI(min_index!1^, min_index!4) 
   sum!2 = 0 // Will lift, when hoisted
   sum!2^ = raise_dim(sum!2, (j:D!0)) 
   for j in range(0, D!0):
      sum!3 = PHI(sum!2^, sum!4)
      d!3 = S!0^ - C!0^
      p!3 = (d!3 * d!3) 
      sum!4 = (sum!3 + p!3)
   sum!3^ = drop_dim(sum!3)     
   !1!2 = (sum!3^ < min_sum!2)
   min_sum!3 = sum!3^
   min_index!3 = i  // Same-level, will lift when hoisted
   min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2)
   min_index!4 = MUX(!1!2, min_index!3, min_index!2)
min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)     
\end{pythonn}
}

@@

Starting at the inner phi-node {\sf sum!3 = PHI(...)}, the algorithm first computes its closure. The closure amounts to the phi-node itself and the addition node {\sf sum!4 = (sum!3 + p!3)}, accounting for the loop-carried dependency of the computation of {\sf sum}. The algorithm replaces this closure with a FOR loop on $j$ removing vectoriaztion on $j$. Note that the SUB and MUL computations remain outside of the loop as they do not depend on phi-nodes that are part of cycles. The dependences are from {\sf p!3[I,J] = (d!3[I,J] * d!3[I,J])} to the monolithic FOR loop and from the FOR loop to {\sf sum!3\^ = drop\_dim(sum!3)}. Lower case index, e.g., \texttt{i}, indicates non-vectorized dimension, while uppercase index, e.g., \texttt{I} indicates vectorized dimension.

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))
for i in range(0, N!0):
  min_sum!2[I] = PHI(min_sum!1^[I], min_sum!4[I]) 
  min_index!2[I] = PHI(min_index!1^[I], min_index!4[I])  
  sum!2 = [0,..,0] 
  sum!2^ = raise_dim(sum!2, (j:D!0))
  d!3[I,J] = S!0^[I,J] - C!0^[I,J]
  p!3[I,J] = (d!3[I,J] * d!3[I,J])
  for j in range(0, D!0):
    sum!3[I,j] = PHI(sum!2^[I,j], sum!4[I,j-1])       
    sum!4[I,j] = (sum!3[I,j] + p!3[I,j])
  sum!3^ = drop_dim(sum!3)     
  !1!2[I] = (sum!3^[I] < min_sum!2[I])
  min_sum!3 = sum!3^
  min_index!3 = i
  min_sum!4[I] = MUX(!1!2[I], min_sum!3[I], min_sum!2[I])
   min_index!4[I] = MUX(!1!2[I], min_index!3[I], min_index!2[I])
min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)
\end{pythonn}
}

When processing the outer loop two closures arise, one for {\sf min\_sum!2[I] = PHI(...)} and one 
for {\sf min\_index!2[I] = PHI(...)}. Since the two closures \emph{do not} intersect, we have two distinct FOR-loops on $i$:

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)
\end{pythonn}
}

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_index!1 = 0
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

// drop_dim becomes redundant
!2!1 = (min_sum!2, min_index!2)
\end{pythonn}
}

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_index!1 = 0
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

sum!2 = [0,..,0]
d!3[I,J] = SUB_SIMD(S!0^[I,J],C!0^[I,J])
p!3[I,J] = MUL_SIMD(d!3[I,J], d!3[I,J])

@@

!2!1 = (min_sum!2, min_index!2)   
\end{pythonn}
}
